---
title: "Exploring Base R's cut() function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring Base R's cut() function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(adcp)
library(dplyr)
library(gbRd)
```

This vignette explores the behavior of base R's `cut()` function, which is used to calculate the intervals in `adcp_count_obs()`. These intervals are used in `adcp_plot_speed_hist()` and can be passed to `adcp_plot_current_rose()` using the `breaks` argument.

```{r, echo=FALSE}
cat("\n")
```


## `cut()` function

First, let's look at the help file for `cut`:
```{r, echo=FALSE, results='asis'}
#cut

Rd <- gbRd::Rd_fun(help("cut")) 
htmlfile <- tempfile(fileext = ".html")
tools::Rd2HTML(Rd, htmlfile, package = "", stages = c("install", "render"))
htmllines <- readLines(htmlfile)
i <- grep("<body>", htmllines)
j <- grep("</body>", htmllines)

cat(htmllines[(i+1):(j-1)], sep = "\n")

```


## Explore with simple example

The arguments we want to explore are `include.lowest`, `right`, and `dig.lab`.

Create a simple dataframe:
```{r}
df <- data.frame(value = seq(1:10))

df
```

### Assign number of intervals

Assign values to 3 even intervals using the default arguments (`include.lowest = FALSE`, `right = TRUE`, and `dig.lab = 3`):
```{r}
df$int1 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = FALSE,
  right = TRUE,
  dig.lab = 3
)

df
```
Notes: 

  - Intervals are **right** inclusive.
  - Outer limits are expanded by 0.1 % of the range (`(10 - 1) * 0.001`).
    - Start of first interval is the minimum value minus 0.1 % of the range.
    - End of the last interval is the maximum value plus 0.1 % of the range.
  - The end of the last interval is really `10.009`. Because `dig.lab = 3`, this is truncated to `10.0`, and `R` drops the trailing zero.

To verify the end of the last interval, set `dig.lab = 5`:
```{r}
df$int2 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = FALSE,
  right = TRUE,
  dig.lab = 5
)

df
```

For intervals that are **left inclusive**, set `right = FALSE`:
```{r}
df$int3 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = FALSE,
  right = FALSE,
  dig.lab = 3
)

df
```

Notes:

  - The assigned interval changes for values of 4 and 7.
  - In this example, `dig.lab = 3`, so the final interval is displayed as `[7, 10)`. The value of `10` is assigned to this right-exclusive interval because it is really `[7, 10.009)`.
  
To verify the last interval set `right = FALSE` and `dig.lab = 5`:
```{r}
df$int4 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = FALSE,
  right = FALSE,
  dig.lab = 5
)

df
```

The behavior of `include.lowest` depends on the value of `right`: 
  - When `right = TRUE`, `include.lowest = TRUE` makes the first interval left-inclusive.
  - When `right = FALSE`, `include.lowest = TRUE` makes the last interval right-inclusive.
  
For `right = TRUE` and `include.lowest = TRUE` (note the two square brackets for the first interval):
```{r}
df$int5 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = TRUE,
  right = TRUE,
  dig.lab = 5
)

df
```

For `right = FALSE` and `include.lowest = TRUE` (note the two square brackets for the last interval):
```{r}
df$int6 <- cut(
  df$value, 
  breaks = 3, 
  include.lowest = TRUE,
  right = FALSE,
  dig.lab = 5
)

df
```


### Assign break values



Start fresh with a simple dataframe:
```{r}
df <- data.frame(value = seq(1:10))

df
```

Assign values to intervals using the breaks `c(1, 4, 7, 10)` using the default arguments (`include.lowest = FALSE`, `right = TRUE`, and `dig.lab = 3`):

```{r}
df$int1 <- cut(
  df$value, 
  breaks = c(1, 4, 7, 10), 
  include.lowest = FALSE,
  right = TRUE,
  dig.lab = 3
)

df
```

Notes:

- The value of `1` is not assigned an interval because `include.lowest = FALSE`.

Set `include.lowest = TRUE` so that the value `1` will be assigned to an interval.

```{r}
df$int2 <- cut(
  df$value, 
  breaks = c(1, 4, 7, 10), 
  include.lowest = TRUE,
  right = TRUE,
  dig.lab = 3
)

df
```

Check to make sure there the break values were not truncated by increasing `dig.lab`"
```{r}
df$int3 <- cut(
  df$value, 
  breaks = c(1, 4, 7, 10), 
  include.lowest = TRUE,
  right = TRUE,
  dig.lab = 5
)

df
```











